---
title: "R Tutorial"
author: |
  | DBMI Group
date: "October 25, 2016"
output:
  beamer_presentation:
    fig_height: 3.5
    fig_width: 5
    keep_tex: no
    toc: yes
---
## R  

R is 'GNU S', a freely available language and environment for statistical computing and graphics which provides a wide variety of statistical and graphical techniques

- linear and nonlinear modeling
- statistical tests
- time series analysis
- classification
- clustering
- genomics


Download and install R (precompiled binary distribution) from the nearest R mirror site, <https://cran.stat.ucla.edu>

## Look and feel of R 

```{r}
1+2
2*2
2*2*2*2
2^4
```

## More calculations in R

```{r}
log(16, 2)
pi
pi*(3^2)
sin(pi/2)
```


## Simple plot in R

```{r}
x = c(-10:10)
y = x^3
plot(x, y)
```


## Make the figure fancier by adding more plotting parameters

```{r}
x = c(-10:10)
y = x^3
plot(x, y, pch=6, lty=3, type="b", col="red")
```


## Try changing parameter values 

```{r}
x = c(-5:10)
y = x^3
plot(x, y, pch=15, lty=2, type="b", col="orange")
```




## R Studio 

RStudio is a set of integrated tools designed to help you be more prudictive with R. It includes 

- console
- syntax-highlighting editor that supports direct code execution
- plotting
- history
- debugging 
- workspace management. 

Download and install RStudio from <https://www.rstudio.com/products/rstudio/download>


## R Studio text editor
R Studio -> File -> New File -> R Script to create an R code file, type commands below in the text file, click the first line, and click Run button three times.

```{r, eval=FALSE}
x = c(-10:10)
y = x^3
plot(x, y, pch=1, lty=1, type="b", col="blue")
```


## R Packages

R packages are collections of functions, data, and complied code in a ready-to-use format with enriched documentation. 

Use R Studio to install R packages 
R Studio -> Tools -> Install Packages... -> type in package name like "ggplot2"

- package name 1 "xlsx"
- package name 2 "foreign"
- package name 3 "ggplot2"


## R Objects

R basic command

- Data input
- Data Management
- Basice Statistics
- Basic Graphs


## Data input

Data input:

- Importing Data  
- Data export  
- Viewing Data  
- Missing Data  


## Importing Data

```{r}
## Keyboard input  
# create a data frame from scratch   
age <- c(25, 30, 56)   
gender <- c("male", "female", "male")  
weight <- c(160, 110, 220)   
mydata <- data.frame(age,gender,weight)  
dim(mydata)
summary(mydata)
str(mydata)

## From web  
# mydata <- read.csv("http://www.ats.ucla.edu/stat/data/hdp.csv")

## From text file  
mydata <- read.table("mydata.txt", header=TRUE, sep="\t")  
    
## From Excel  
library(xlsx)  
# read in the worksheet named mysheet  
mydata <- read.xlsx("/Users/hai/Dropbox (DBMI)/Rtutorial/testdata.xlsx", sheetName = "mysheet")  
```

## Data export

```{r}
# To A Tab Delimited Text File
write.table(mydata, "/Users/hai/Dropbox (DBMI)/Rtutorial/mydata.txt", sep="\t")


# To an Excel Spreadsheet
library(xlsx)
# write.xlsx(mydata, "/Users/hai/Dropbox (DBMI)/Rtutorial/mydata.xlsx")
```


## Viewing Data

```{r}
# list objects in the working environment
ls()

# list the variables in mydata
names(mydata)

# list the structure of mydata
str(mydata)

# print mydata 
mydata

# print first 10 rows of mydata
head(mydata, n=10)

# print last 5 rows of mydata
tail(mydata, n=5)
```


## Missing Data

```{r}
# Testing for Missing Values
x <- c(1,2,NA,3)
is.na(x) 

y <- c(1,2,3,NA)
is.na(y) 

# Excluding Missing Values from Analyses

mean(x) # returns NA
mean(x, na.rm=TRUE) # returns 2

# list rows of data that have missing values 
mydata$x <- x
mydata$y <- y
mydata[!complete.cases(mydata),]

# create new dataset without missing data 
newdata <- na.omit(mydata)
# rm(list=ls())

```

## Data Management

Data Management:

- Creating new variables  
- Control Structures  
- Sorting Data and Merging Data  
- Subsetting Data  
- Using with( ) and by( )  


## Creating new variables

```{r}
# Creating new variables
mydata$x <- c(1,2,NA,3)
mydata$y <- c(1,2,3,NA)
mydata$sum <- mydata$x + mydata$y
mydata$mean <- (mydata$x + mydata$y)/2

# Recoding variables
# create 2 age categories 
mydata$age <- c(37,49,79,59,49,90,37,49,79,59,49,90,37,49,79,59,49,90,45,23)
mydata$agecat <- ifelse(mydata$age > 70, 
c("older"), c("younger")) 
```

## Control Structures

- if-else  
if (cond) expr  
if (cond) expr1 else expr2  

- for  
for (var in seq) expr  

- while  
while (cond) expr  

- switch  
switch(expr, ...)  

- ifelse  
ifelse(test,yes,no)  


## Sorting Data and Merging Data

```{r}
# sort by mpg and cyl
attach(mtcars)
newdata <- mtcars[order(mtcars$mpg, mtcars$cyl),]

# Adding Columns
# merge two data frames by ID and Country
mtcars$ID <- rownames(mtcars)
data_frameA <- mtcars[,c(1,2,3,4,5,6,12)]
data_frameB <- mtcars[,c(6,7,8,9,10,11,12)]
total <- merge(data_frameA,data_frameB,by=c("ID","wt"))

# Adding Rows
data_frameA <- head(mtcars)
data_frameB <- tail(mtcars)
total <- rbind(data_frameA, data_frameB)
```

## Subsetting Data part 1
```{r}
# Selecting (Keeping) Variables

#select by variable names "mpg", "cyl", "disp"
mydata <- mtcars
myvars <- c("mpg", "cyl", "disp")
newdata <- mydata[myvars]

# select 1st and 5th thru 10th variables
newdata <- mydata[c(1,5:10)]

# Excluding (DROPPING) Variables
#exclude variables "mpg", "cyl", "disp"
mydata <- mtcars
myvars <- names(mydata) %in% c("mpg", "cyl", "disp") 
newdata <- mydata[!myvars]

# exclude 3rd and 5th variable 
newdata <- mydata[c(-3,-5)]
```

## Subsetting Data part 1
```{r}
# Selecting Observations
#first 5 observations
newdata <- mydata[1:5,]

# based on variable values
newdata <- mydata[ which(gear== 3 & wt > 65),]

# Selection using the Subset Function
#using subset function 
newdata <- subset(mydata, mpg >= 20 & mpg < 30, 
select=c(ID, wt))
```

## Using with( ) and by( )

- With  
with(data, expression)  
example applying a t-test to a data frame mydata  
with(mydata, t.test(y ~ group))  

- By  
by(data, factorlist, function)  
example obtain variable means separately for  
each level of byvar in data frame mydata   
by(mydata, mydata$byvar, function(x) mean(x))  


## Basice Statistics

Basice Statistics   
- t-test & Correlations  
- ANOVA   
- Multiple Regression   
- Regression Diagnostics   


## t-test & Correlations

- Correlations  
```{r}
# Correlation matrix from mtcars
# with mpg, cyl, and disp as rows 
# and hp, drat, and wt as columns 
x <- mtcars[1:3]
y <- mtcars[4:6]
cor(x, y)
```
- t-tests
```{r}
# independent 2-group t-test
t.test(mtcars$mpg, mtcars$wt) 
# one sample t-test
t.test(mtcars$mpg,mu=20) # Ho: mu=20
```

## ANOVA

```{r}
# Analysis of variance  
fit <- aov(mpg ~ wt + cyl, data=mtcars)
fit
```

## Multiple Regression

```{r}
## Fitting the Model

# Multiple Linear Regression Example 
fit <- lm(mpg ~ wt + cyl + hp, data=mtcars)
summary(fit) # show results
residuals(fit) # residuals
anova(fit) # anova table 

# Diagnostic Plots
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(fit)
par(mfrow=c(1,1)) # back to 1 graphs/page

# Comparing Models
fit1 <- lm(mpg ~ wt + cyl + hp, data=mtcars)
fit2 <- lm(mpg ~ wt + cyl, data=mtcars)
anova(fit1, fit2)
```


## Multiple Regression (cont)

```{r}
## Variable Selection
#Stepwise Regression
library(MASS)
fit <- lm(mpg ~ wt + cyl + hp, data=mtcars)
step <- stepAIC(fit, direction="both")
step$anova # display results

# All Subsets Regression
library(leaps)
leaps<-regsubsets(mpg ~ wt + cyl + hp,data=mtcars,nbest=10)

# view results 
summary(leaps)

# plot a table of models showing variables in each model.
# models are ordered by the selection statistic.
###plot(leaps,scale="r2")

# plot statistic by subset size 
library(car)
###subsets(leaps, statistic="rsq")
```

## Regression Diagnostics

```{r}
## Assume that we are fitting a multiple linear regression

library(car)
fit <- lm(mpg~disp+hp+wt+drat, data=mtcars)
```
## Outliers
```{r}
# Assessing Outliers
outlierTest(fit) # Bonferonni p-value for most extreme obs
qqPlot(fit, main="QQ Plot") #qq plot for studentized resid 
```
## Outliers (cont)
```{r}
leveragePlots(fit) # leverage plots
```

## Influential Observations
```{r}
# added variable plots 
###av.Plots(fit) 
```

## Cook's D plot
```{r}
# identify D values > 4/(n-k-1) 
cutoff <- 4/((nrow(mtcars)-length(fit$coefficients)-2)) 
plot(fit, which=4, cook.levels=cutoff)
```
## Influence Plot 
```{r}
influencePlot(fit,  id.method="identify", main="Influence Plot", sub="Circle size is proportial to Cook's Distance" )
```

## Non-normality
```{r}
# Normality of Residuals
# qq plot for studentized resid
qqPlot(fit, main="QQ Plot")
```

## distribution of studentized residuals
```{r}
library(MASS)
sresid <- studres(fit) 
hist(sresid, freq=FALSE, 
   main="Distribution of Studentized Residuals")  
xfit<-seq(min(sresid),max(sresid),length=40) 
yfit<-dnorm(xfit) 
lines(xfit, yfit)
```

## Non-constant Error Variance
```{r}
# Evaluate homoscedasticity
# non-constant error variance test
ncvTest(fit)
```

## plot studentized residuals vs. fitted values 
```{r}
spreadLevelPlot(fit)
```

## Multi-collinearity
```{r}
# Evaluate Collinearity
vif(fit) # variance inflation factors 
sqrt(vif(fit)) > 2 
```

## Nonlinearity
```{r}
# Evaluate Nonlinearity
# component + residual plot 
crPlots(fit)
```

## Ceres plots 
```{r}
ceresPlots(fit)
```

## Non-independence of Errors
```{r}
# Test for Autocorrelated Errors
durbinWatsonTest(fit)
```

## Basic Graphs

Basic Graphs:  
- Creating a Graph  
- Density Plots  
- Dot Plots  
- Bar Plots  
- Boxplots  
- Scatter Plots  


## Creating a Graph

```{r}
with(mtcars, plot(wt, mpg) )
abline(lm(mpg~wt))
title("Regression of MPG on Weight")
```

## Density Plots
```{r}
hist(mtcars$mpg)
```

## Dot Plots
```{r}
dotchart(mtcars$mpg,labels=row.names(mtcars),cex=.7,
    main="Gas Milage for Car Models", 
   xlab="Miles Per Gallon")
```

## Bar Plots
```{r}
counts <- table(mtcars$gear)
barplot(counts, main="Car Distribution", 
    xlab="Number of Gears")
```    
    
## Boxplots
```{r}
# Boxplot of MPG by Car Cylinders 
boxplot(mpg~cyl,data=mtcars, main="Car Milage Data", 
    xlab="Number of Cylinders", ylab="Miles Per Gallon")
```    
    
## Scatter Plots
```{r}
plot(wt, mpg, 
    xlab="Car Weight ", ylab="Miles Per Gallon ", pch=19)
#Add fit lines
abline(lm(mpg~wt), col="red") # regression line (y~x) 
lines(lowess(wt,mpg), col="blue") # lowess line (x,y)
```

## Scatterplot Matrices
```{r}
pairs(~mpg+disp+drat+wt,data=mtcars)
```